---
title: "Course Work 2"
author: "Hamdha, Hemara, Lakshika"
date: "2026-03-05"
output: html_document
---

United States Advance Retail Sales: Retail Trade
```{r}
library(ggplot2)
library(forecast)
library(tseries)
```
1.DATA PREPARATION AND VISUALIZATION
```{r}
sales_data <- read.csv("C:/Users/hamdh/Downloads/RSXFSN.csv")
head(sales_data)
```

```{r}
# Convert Data Column
sales_data$observation_date <- as.Date(sales_data$observation_date)
# Change the data type to Time series(ts)
retails_ts <- ts(sales_data$RSXFSN, c(2010,12), frequency = 12)
class(retails_ts)
```

```{r}
#library(ggplot2)

# Plot Raw data(Sales_data)
plot(retails_ts, col = "blue", lwd = 2, 
     main = "U.S. Monthly Retail Sales (2010-2025)", 
     ylab = "Sales (Millions of $)", xlab = "Year")

```

```{r}
# Seasonal Decomposition
decomp <- decompose(retails_ts)
plot(decomp)
```

```{r}
# Log Transformation (To stabilize the Variance)
log_retails_ts<- log(retails_ts)
```

```{r}
adf.test(log_retails_ts)
```

2.STATIONARY AND DIFFERENCING
```{r}
#library(forecast)


# Seasonal Differencing (D=1) to remove seasonal patterns
diff1_12 <- diff(log_retails_ts, lag = 12)
plot(diff1_12, main="Seasonally Differenced Log Data")

```

```{r}
#  First Differencing (d=1) to remove remaining trend
stationary_ts <- diff(diff1_12, lag = 1)
plot(stationary_ts,main="Fully Differenced stationary Data")
```

```{r}
# library(tseries)
#  Check stationarity (Looking for p-value < 0.05) to confirm stationary
adf.test(stationary_ts)
```


```{r}
# Check ACF/PACF (Model Identification)

par(mfrow=c(1,1))
acf(stationary_ts, main="ACF (Identification)")
# seasonal spike at Lag 12 this confirms that even after differencing, there is a strong seasonal relationship.It justifies why the (auto.arima) chose a Seasonal Moving Average (SMA) component.
```

```{r}
pacf(stationary_ts, main="PACF (Identification)")
par(mfrow=c(1,1))
# Early significant PACF spikes indicate the presence of non-seasonal AR terms
# This helps identify the AR component of the SARIMA model
   
```

```{r}
end(retails_ts)
```


3.MODEL VALIDATION
```{r}
# Data Split & Validation(MAPE)
# Split: Train(2020-2024) and Test(2025)
train_data <- window(log_retails_ts, end = c(2024, 12))
test_data  <- window(log_retails_ts, start = c(2025, 1))

# Fit model on Training set
fit_val <- auto.arima(train_data, seasonal = TRUE, 
                      stepwise = FALSE, approximation = FALSE)

# Forecast the test period (12 months)
forecast_val <- forecast(fit_val, h = 12)

# Calculate MAPE on Original Scale (Back-transform with exp)
actual_test_original <- exp(as.numeric(test_data))
predicted_test_original <- exp(as.numeric(forecast_val$mean))

mape_val <- mean(abs((actual_test_original - predicted_test_original) / actual_test_original)) * 100
cat("\nValidation MAPE (Original Scale):", round(mape_val, 2), "%\n")


```


```{r}
# Validation Plot :Red line (Actual) should follow the Blue line (Predicted)
# This creates the blue forecast line and the gray confidence intervals
plot(forecast_val, 
     main="Model Validation: Actual vs Predicted (2025)", 
     xlab="Year", 
     ylab="Log Retail Sales",
     lwd=1)

# 2. Add the ACTUAL test data (The red line)
# Note: Since your forecast is on log scale, test_data must also be log
lines(test_data, col="red", lwd=2)

# 3. Add the legend to explain the colors
legend("topleft", 
       legend=c("Forecasted (Train)", "Actual (Test)"), 
       col=c("blue", "red"), 
       lty=1, 
       lwd=2)
```
4. FINAL MODELING AND FORECASTING
```{r}
# Model Fitting
# Using auto.arima to find the best SARIMA (p,d,q)(P,D,Q) model
fit_sarima <- auto.arima(log_retails_ts, seasonal = TRUE, 
                         stepwise = FALSE, approximation = FALSE)

# Summary of the fitted model
summary(fit_sarima)
```
The auto.arima function identified the optimal model as SARIMA(1,0,2)(0,1,1)[12] with drift. The high AR1 coefficient (0.9692) indicates a strong month-to-month dependency, while the significant seasonal moving average term (SMA1 = -0.8255) confirms the presence of consistent yearly patterns. The drift term of 0.0037 accounts for the overall growth trend in the U.S. retail sector. With a remarkably low Training MAPE of 0.14% and a highly negative AIC, the model demonstrates a superior fit, making it a reliable tool for future forecasting.


```{r}
#Residual Diagnostic Analysis (checking white noise)
residuals <- residuals(fit_sarima)

# Perform the Ljung-Box test on the residuals(p-value>0.05)
ljung_box_test<- Box.test(residuals(fit_sarima), lag = 24, type = "Ljung-Box")
ljung_box_test
```
this is significantly greater than the 0.05 threshold, we fail to reject the null hypothesis. This statistically proves that our residuals are independent and behave like White Noise.

```{r}
# 2. Advanced Normality Diagnostics (Histogram and Q-Q Plot)
# These prove that your model residuals are "White Noise"
par(mfrow = c(1, 2)) # Put plots side-by-side
# Histogram of Residuals
hist(residuals, breaks = 20, main = "Histogram of Residuals", 
     col = "lightblue", border = "black", xlab = "Residuals")
# Q-Q Plot
qqnorm(residuals, main = "Normal Q-Q Plot")
qqline(residuals, col = "red", lwd = 2)
par(mfrow = c(1, 1)) # Reset plot layout
```

```{r}
# 4. FINAL FORECASTING

# 1. Generate the 24-month forecast
forecast_result <- forecast(fit_sarima, h = 24)

# 2. Plot the Forecast in Original Scale
# We use exp() on the entire forecast object to keep the time-axis (Years)
plot(exp(forecast_result$mean), 
     type="l", 
     col="blue", 
     lwd=2, 
     main="24-Month Forecast of Retail Sales (Original Scale)", 
     ylab="Retail Sales (Million USD)", 
     xlab="Year")

# 3. Create the Business-Ready Table (Back-transformed to USD)
forecast_df <- as.data.frame(forecast_result)

# Note on Indexing: 
# Column 1 = Point Forecast
# Column 2 = Lo 80 | Column 3 = Hi 80
# Column 4 = Lo 95 | Column 5 = Hi 95
final_report_table <- data.frame(
  Month           = rownames(forecast_df),
  Predicted_Sales = round(exp(forecast_df[, 1]), 2), # Point Forecast
  Lower_Limit_95  = round(exp(forecast_df[, 4]), 2), # Corrected: Lo 95
  Upper_Limit_95  = round(exp(forecast_df[, 5]), 2)  # Corrected: Hi 95
)

# 4. View the final results
print(final_report_table)
```




